---
title: "Tutorial for main functions in SurrogateRank package"
output:
  html_document:
editor_options: 
  chunk_output_type: console
---

```{r global_options, echo=FALSE}
knitr::opts_chunk$set(fig.width=13, warning=FALSE, tidy=TRUE, tidy.opts=list(width.cutoff=60))
```

In this tutorial, we will go through examples using the functions in the *SurrogateRank* package. First, install the package from CRAN and load it.

```{r results = "hide", message=FALSE}
#install.packages("SurrogateRank")
library(SurrogateRank)

# And some other useful packages
library(dplyr)
```

This package uses rank-based nonparametric approaches to evaluate surrogate markers in a small sample size setting. 

This approach, intended for use in the single-surrogate, single-trial context was described in Parast L, Cai T, & Tian L (2024). A rank-based approach to evaluate a surrogate marker in a small sample setting. Biometrics, 80(1), ujad035. 

In addition, these approaches were adapted to the high-dimensional setting, as described in Hughes A., Parast L., Thiébaut R., Hejblum B.P. (2025). RISE : Two-Stage Rank-Based Identification of High-Dimensional Surrogate Markers Applied to Vaccinology. Statistics in Medicine, 44(20-22): e70241. 

Finally, these approaches were extended to the multi-trial setting, as described in Hughes A., Thiébaut R., Hejblum B.P. (2026). Meta-Analysis of High-Dimensional Surrogate Markers (manuscript submitted).

We describe each of these three methods below.

If you are viewing the html version of this tutorial, you can find the R markdown file here: <https://github.com/laylaparast/SurrogateRank>. It is the file called SurrogateRank_tutorial.Rmd.

# Single Surrogate: Evaluation and Power 

Let's first take a look at the example dataset (for one surrogate marker) included in the package, *example.data*. The dataset has 25 observations from a treatment group (group 1) and 25 observations from a control group (group 0), and contains *y1*, the primary outcome in the treatment group, *y0*, the primary outcome in the control group, *s1*, the surrogate marker in the treatment group, and *s0*, the surrogate marker in the control group.

```{r}
data(example.data)
names(example.data)
```

Now, we calculate the metric $\delta$ which measures the strength of the surrogate using the function *delta.calculate*. This function requires the primary outcome and surrogate marker in each group, as shown below. The function returns the U-statistic for the treatment effect on the primary outcome, *u.y*, the U-statistic for the treatment effect on the surrogate marker, *u.s*, the difference between these, *delta.estimate*, and the standard error of each estimate, *sd.u.y, sd.u.s, sd.delta*. Lower values of *delta.estimate* indicate a stronger surrogate marker in the sense that the treatment effect on the surrogate is close to the treatment effect on the primary outcome.

```{r}
delta.calculate(yone = example.data$y1, yzero = example.data$y0, sone = example.data$s1, szero = example.data$s0)
```

Next, we formally test this *delta* using *test.surrogate*. The required arguments for this function are the same as *delta.calculate* In addition to the same returned quantities as *delta.calculate*, this function also returns the 1-sided confidence interval for *delta*, the epsilon that is used for the surrogacy test, and an indication of whether the provided surrogate is a valid surrogate based on this testing approach. For example, with this dataset, *is.surrogate* is *TRUE* meaning we can conclude this surrogate is valid.

```{r}
test.surrogate(yone = example.data$y1, yzero = example.data$y0, sone = example.data$s1, szero = example.data$s0)
```

Optional arguments that *test.surrogate* takes are 1) *epsilon* which is the threshold to use for testing delta where the default calculates epsilon as a function of desired power for testing for a treatment effect using the surrogate, 2) *power.want.s* which is the desired power in (1), used to calculate *epsilon* if it is not provided; the default is 0.7, 3) *u.y.hyp* which is the hypothesized value of u.y used in the calculation of epsilon; the default is to use the estimated value *u.y*. For example, in the code below I specify the epsilon, making it more strict than what is used by default above.

```{r}
test.surrogate(yone = example.data$y1, yzero = example.data$y0, sone = example.data$s1, szero = example.data$s0, epsilon = 0.1)
```

Next, the function *est.power* calculates the estimated power to detect a valid surrogate given a total sample size and specified alternative. This function would generally be used in a design stage before data are collected. The arguments are: 1) *n.total*, the total sample size in the study, 2) *rho*, the rank correlation between Y and S in group 0, default is 0.8, 3) *u.y.alt*, the specified alternative for *u.y*, 4) *delta.alt*, the specified alternative for *delta*, 5) *power.want.s*, the desired power for *u.s*, default is 0.7. The function returns the estimated power to detect a valid surrogate, given these specifications. For example, let's say we are planning a study with a sample size of 50 and based on prior data or domain knowledge we specify the rank correlation between Y and S in group 0 to be 0.7, the alternative for *u.y* to be 0.9, and the specified alternative for *delta* to be 0.05, and our desired power for *u.s* is 0.6. Then, our estimated power is 0.84.

```{r}
est.power(n.total = 50, rho = 0.7, u.y.alt=0.9, delta.alt = 0.05, power.want.s=0.6)
```

# RISE: Rank-Based Identification of High-Dimensional Surrogate Markers

In this section, we will illustrate the functions for a high-dimensional surrogate. Again, let's start by looking at the example dataset (for the high-dimensional setting) in the package.

```{r}
data(example.data.highdim)
names(example.data.highdim)
```

As before, this data contains 25 observations from a treatment group (group 1) and 25 observations from a control group (group 0). As in the other example, *y1* and *y0* are the values of the primary response in the treated and control group, respectively. However, this time, *s1* and *s0* are matrices of 1000 candidate surrogate markers in the treated and control group, respectively. The final element of the data, *hyp*, gives whether the surrogate candidate is valid or not, where 'null false' corresponds to a valid surrogate candidate and 'null true' an invalid surrogate. Of course, we only have this variable because this is simulated data and in practice, we would not have this information; we include it here for illustration. 

Let's first take a look at the *hyp* variable to see which of the 1000 markers are true positives that we hope to be identified by RISE, and which are not. 

```{r}
which(example.data.highdim[["hyp"]] == 'null false')
```

So, the first 100 of the simulated markers are true positives, and the remaining 900 are negatives. Let's keep this in mind for later. 


RISE works in two stages, which are generally applied on different (random) splits of the data in order to avoid overfitting. The first stage, called screening, allows for a selection of strong candidate surrogates by applying the aforementioned univariate test to each candidate, then correcting the p-values for the high multiplicity of tests. Let's perform the screening stage using the *rise.screen()* function on the example data (without sample splitting).

```{r}
screen.result = rise.screen(yone = example.data.highdim$y1,
            yzero = example.data.highdim$y0,
            sone = example.data.highdim$s1,
            szero = example.data.highdim$s0,
            power.want.s = 0.8)
```

Let's check the outputs of this function. The *screening.metrics* element returns a data frame with results from the testing of each marker one-by-one.

```{r}
screen.result[["screening.metrics"]] %>% head()
```

The next element *significant.markers* gives the names of the markers which were judged significant at the provided $\alpha$ level (0.05 by default) after test multiplicity correction. Now, remember that markers 1 through 100 were true positives. Let's check to see how many of our identified positives were true. 

```{r}
screen.result[["significant.markers"]]
```

All of the claimed positives were truly positive, with no false positives. How many markers in total were determined significant?

```{r}
screen.result[["significant.markers"]] %>% length()
```

So, 18/100 true positives were not identified. This reflects the fact that the non-parametric methods employed in RISE tend to make it *conservative*. That is, it controls the false positive rate well but may be less powered than parametric methods (when those parametric assumptions hold!).


The final element is a data frame giving the weights to pass to the evaluation stage to form a synthetic marker as a weighted, standardised sum of the markers screened as significant in the screening stage. We give a couple of different options for how to specify these weights. The default option, specified by setting *weight.mode = "inverse.delta"* in *rise.screen()*, sets the weights as the inverse of the estimated delta values. Let's have a look at the weights given by the function

```{r}
screen.result[["screening.weights"]] %>% head()
```

The *weight_raw* column gives the pre-normalised weights (i.e. simply the inverse of the delta values). The *weight* column divides these weights by the maximum weight, such that the weights now represent a proportion of the maximum value. This is simply for easier interpretation, as using the raw or standardised weights will not change the results of the evaluation. This is because RISE is invariant to *monotonic* (i.e. rank-preserving) transformations. 

Now let's perform the evaluation stage using the *rise.evaluate()* function. For simplicity, let's evaluate on the same data on which we screened. Although, in practice, you should probably perform sample splitting on the data to avoid overfitting bias. Note that we have to pass some elements from the screening stage into this function (the names of the significant markers and the weights).

```{r}
evaluate.results = rise.evaluate(yone = example.data.highdim$y1,
            yzero = example.data.highdim$y0,
            sone = example.data.highdim$s1,
            szero = example.data.highdim$s0,
            power.want.s = 0.8,
            markers = screen.result[["significant.markers"]],
            screening.weights = screen.result[["screening.weights"]])
```

Let's look at the elements of this output. The first is *individual.metrics*, which is basically a re-application of the screening phase in the evaluation data. This can be a useful tool for assessing overfitting - if none of the markers which were significant in the screening phase were significant in the evaluation data, there is probably some overfitting going on. Since, in this example, we used exactly the same data to screen and evaluate, this will give us the same results as before (with the adjusted p-values slightly different due to the different number of tests).

```{r}
evaluate.results[["individual.metrics"]] %>% head()
```

The second element, *gamma.s* gives us the values of the composite marker $\gamma$ for the treated and untreated groups.

```{r}
evaluate.results[["gamma.s"]][["gamma.s.one"]] %>% head() # treated gamma
evaluate.results[["gamma.s"]][["gamma.s.zero"]] %>% head()# untreated gamma
```

We see that the values of the composite surrogate are large and positive in the treated group, and large and negative in the untreated group, suggesting good separation of treatment groups by this variable. We can see that more formally in the third element *gamma.s.evaluate*, which gives us the results of the testing of gamma as a univariate surrogate in the evaluation data.

```{r}
evaluate.results[["gamma.s.evaluate"]]
```

The small delta value and p-value provide us evidence that this composite marker is a strong trial-level surrogate in this data. 

The final element, *gamma.s.plot*, gives us a plot of the rankings of the composite marker gamma.s against the response in the evaluation data.

```{r}
evaluate.results[["gamma.s.plot"]]
```

If gamma is a strong trial-level surrogate, we expect separation of the treatment groups by gamma to be similar to that by the primary outcome. In this case, this is what we see, and this is made explicit by the annotation of the value of delta for the combined marker, which is close to 0. We may also have high spearman correlation between the *values* of the primary response and gamma (i.e. the points follow the diagonal line closely). A high level of correlation on the individual-level is a property of what is referred to in the literature as *individual-level surrogate markers*. While RISE is a method made for identifying *trial-level surrogate markers*, it is still interesting to see where trial- and individual-level properties may co-occur.

We can also apply both stages of RISE together with the function *test.surrogate.rise()*. This gives us two elements, *screening.results* and *evaluation.results* which are exactly the outputs of the individual screening and evaluation functions as explained above. Note that this function automatically performs random sample splitting on the data, by default at ratio screening:evaluation of 2:1. This ratio is controlled by the *screen.proportion* argument. If this is set to 1, the screening and evaluation are performed on the same dataset.

```{r}
test.surrogate.rise.result = test.surrogate.rise(yone = example.data.highdim$y1,
            yzero = example.data.highdim$y0,
            sone = example.data.highdim$s1,
            szero = example.data.highdim$s0,
            power.want.s = 0.8,
            screen.proportion = 2/3)
```

Let's check the results from this sample-split application of RISE. How many significant markers were identified?

```{r}
test.surrogate.rise.result[["screening.results"]][["significant.markers"]] %>% length()
```

Now, what were the resulting metrics and plot in the 1/3 of the data left for evaluation?

```{r}
test.surrogate.rise.result[["evaluation.results"]][["gamma.s.evaluate"]]
test.surrogate.rise.result[["evaluation.results"]][["gamma.s.plot"]]
```

We also see good separation of the treatment groups on the plot, corresponding to small values of delta and the p-value, indicating that RISE identified a strong combined trial-level surrogate, evaluated on independent data. 

# RISE-Meta: Meta-Analysis of High-Dimensional Surrogate Markers

Finally, we will describe the extension to the multi-trial setting. This is applicable when we have access to data concerning the same treatment, candidate markers, and primary endpoint, across multiple trials. 

The idea behind this method is similar to the above approaches. First, the RISE method is applied to the patient level data to derive point and standard error estimates for the surrogacy metric $\delta$ for each marker, within each distinct study. Then, these estimates are pooled across studies using a parametric meta-analysis model. Hypothesis testing is used to derive a set of promising markers. Finally, these markers are combined together to form a composite signature, which may then be re-evaluated (ideally on unseen data). 

Let's start by loading the example data. 

```{r}
data(example.data.highdim.multistudy.ipd)
names(example.data.highdim.multistudy.ipd)
```

The form of this data is very similar to what we had in the previous section. We have *y1* and *y0* describing the values of the primary endpoint in the treated and control groups, matrices *s1* and *s0* whose columns are surrogate candidates, and *hyp* giving us the truth of the validity of the surrogates. However, we also have *study1* and *study0*, giving us the names of the studies corresponding to each sample. Let's take a look at these: 

```{r}
head(example.data.highdim.multistudy.ipd[["study1"]])
head(example.data.highdim.multistudy.ipd[["study0"]])

table(example.data.highdim.multistudy.ipd[["study1"]])
table(example.data.highdim.multistudy.ipd[["study0"]])
```

For convinience, these are arranged such that all the participants from the same study are next to each other. There are five studies in total, each with 25 pariticipants in each arm. 

The main functions for RISE-Meta are *rise.screen.meta()* and *rise.evaluate.meta()*. 

*rise.screen.meta()* applies RISE within each study and then combines the surrogacy estimates $\delta$ using meta-analysis. Let's see how this works with our example data: 

```{r}
rise.meta.screen.result <- rise.screen.meta(
  yone = example.data.highdim.multistudy.ipd$y1,
  yzero = example.data.highdim.multistudy.ipd$y0,
  sone = example.data.highdim.multistudy.ipd$s1,
  szero = example.data.highdim.multistudy.ipd$s0,
  studyone = example.data.highdim.multistudy.ipd$study1,
  studyzero = example.data.highdim.multistudy.ipd$study0,
  epsilon.study = 0.2,
  epsilon.meta = 0.2,
  meta.analysis.method = "RE",
  n.cores = 1
)
```

Before looking at the results, let's explain some of the important arguments. *epsilon.study* corresponds to the value of $\epsilon$ for the within-study analyses, while *epsilon.meta* gives the value of $\epsilon$ for the meta-analysis. In this example, we set these ourselves to a common value. *meta.analysis.method* gives the requested method for meta-analysis, which includes "RE" (random-effects), where the within-study effects are assumed to come from a common distribution, and "FE" (fixed-effect), where the within-study effects are assumed to be estimating the same common effect. We recommend using the random-effects model. Finally, the *n.cores* corresponds to the number of cores to use when doing parallel computing. Here, we set this to 1 (i.e. no parallel computation), but this can speed up the computation when datasets are larger that this example. 


Now let's examine some of the outputs. 

```{r}
names(rise.meta.screen.result)
```

*screening.metrics.study* gives the within-study results. Effectively, this gives the results from the RISE method applied to each study like above, before considering pooling effects across studies. For example, let's check the first marker evaluated in the first study

```{r}
rise.meta.screen.result[["screening.metrics.study"]][1,]
```
Here, we see that marker1 has a small $\delta$ value and is highly significant within this study. 

*screening.metrics.meta* gives the results of the meta-analysis for each marker across studies. Again, let's check the first marker in the data

```{r}
rise.meta.screen.result[["screening.metrics.meta"]][1,]
```

Here, we *mu.delta* gives the pooled effect $\mu$ and its confidence interval *ci.delta.lower* and *ci.delta.upper* across the 5 studies. We see that this is close to 0, indicating strong surrogacy across studies. *pi.delta.lower* and *pi.delta.upper* give the bounds for the *prediction interval*, which give the likely range of effects for new, similar studies. *tau2* gives the estimate of $\tau^{2}$, the parameter in the meta-analysis model which quantifies the between-study heterogeneity. *I2* gives the estimate of the $I^{2}$ statistic, which is the percentage of variability which can be attributed to differences across studies. The greater this is, the greater the total variability is dominated by between-study differences and not within-study differences. The *ccc* gives the *concordance correlation coefficient* between the study-level treatment effects on the primary endpoint and on the surrogate candidate. In this case, it is relatively low, indicating only moderate concordance. *p.unadjusted* and *p.adjusted* give the unadjusted and adjusted p-values resulting from the equivalence test for surrogacy. We see that this is very small, indicating high evidence that $\mu$ is close enough to zero relative to the equivalence bound $\epsilon$ to conclude valid surrogacy.

*significant.markers* is a vector giving the significant markers after multiplicity corrections. 

```{r}
rise.meta.screen.result[["significant.markers"]]
```

Here, we identified 9 out of the 10 truly positive markers. 

*screening.weights* is a dataframe giving the relative weights to be used in the next stage for each marker.

```{r}
rise.meta.screen.result[["screening.weights"]] %>%
  arrange(desc(weight)) %>% 
  head()
```

We see that marker 5 has the highest possible weight at 1 and the others are given lower relative weights. 

Finally, we can access various visualisations of these results through the *gamma.s.plot* object. 

Firstly, we can visualise the similarity between the significant markers identified within each study through the *similarity.plots* object. This has two sub-plots. The first is a classic venn diagram showing the similarity between the markers identified by each study. 

```{r}
rise.meta.screen.result[["gamma.s.plot"]]$similarity.plots$venn.plot
```

The second is an upset plot, which is a modification of the venn diagram which is useful when the number of studies becomes large. 

```{r}
rise.meta.screen.result[["gamma.s.plot"]]$similarity.plots$upset.plot
```

Here we see that the within-study analysis identify many common markers. 

The main plot for the screening stage to visualise the results of the marker-level meta-analyses is the *screen.plot*

```{r}
rise.meta.screen.result[["gamma.s.plot"]][["screen.plot"]]
```

By default, this shows the top 15 markers by p-value, although this can be changed in *rise.screen.meta()* through the *screen.plot.topN* argument. For these top markers, it gives the pooled effect $\mu$ for each one. The colour of the points corresponds to the raw p-values from the equivalence testing. 

That's all for now!

------------------------------------------------------------------------
